|

### 2020-11-24 paper/v1/broadband.speed.covid.Rmd

During the pandemic, working from home using digital technologies, whether partially 
or exclusively, was transformed from a niche means of accessing work, albeit one that 
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs
and, to a varying extent, maintained productivity during periods of strict
lockdown around the world.
However, this ability has not been evenly distributed socially or spatially,
creating a new type of digital divide.
On one side are those who can work from home, supported by digital technologies, 
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this new digital divide can be viewed as mainly occupational, 
we consider whether the divide is also technological.
Using the UK as a case study, this paper aims to understand how the resilience
of broadband speeds in terms of both quality and reliability of service may reinforce
or redress the spatial and social dimensions of the digital division exposed by the
pandemic.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK. **need to add in urban-rural**
Our analysis enables us to better understand how the spatial and social distribution 
of both occupations and online accessibility intersect to enable or hinder the 
practice of telecommuting at a time of extreme demand. 
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a common means of accessing work, and broadband
services and infrastructure must be fit for purpose.
**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**

<!-- Our framework: capability to WFH; HB old PARA2 -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of 
location or co-location with colleagues, including supervisors 
[@allen2015effective; @singh2013modeling].
However, successful telecommuting also requires that the quality and
reliability of digital services, particularly home internet connection speeds,
enable the completion of work tasks with a minimum of delay or interruption.
Prior to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and connecting to
colleagues and workplace resources via broadband was the purview of a small minority 
of workers. 
Instead, in the UK, for example, benchmarking broadband performance and service 
delivery by different Internet Service Providers (ISPs) focused upon the quality of 
service for evening leisure use, when video streaming services are at their peak, 
rather than during the working day [@ofcom2017].
Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than the pre-Covid, raising questions around whether 
internet services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked soley from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020; 
@ons2020lm2019]. 
A back of the envelope calculation suggests that up to 40% of the working force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries around the world
[@felstead2020homeworking].
For instance, $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid].
Furthermore, a recent estimate indicated that $37$% of all jobs in the US can be
permanently performed entirely from home [@NBERw26948].

<!-- ICT infrastructure and digital divides-->
None of these changes could have happened in the absence of reliable ICT
infrastructure -- both in terms of software and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature describes first level digital divides in terms of the availability and
quality of internet connectivity, such as that manifest in different geographies 
in the UK [@riddlesden2014broadband; @philip2017digital]. 
Second level digital divides consider the presence or lack of the necessary skills 
to effectively utilise digital technologies and the internet 
[@blank2014dimensions; @van2011internet].
The third level focuses on the differentiated returns of internet use **for...?**
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic 
resilience of people and places during the current pandemic.
We identify this as a new digital divide, one that fundamentally alters the 
potential returns of internet use for the user and wider community, 
assumes skills that are present in some occupations but not in others, 
and relies upon access to high quality internet services.
As the quality of internet infrastructure and services, as well as variation in 
occupations are spatially dependent and clustered in space, our approach offers a 
framework for understanding which types of places, socially and spatially, 
are more likely to land on the right side of this new digital divide. 
By asking how resilient broadband speeds are as experienced in different parts of the 
UK during a time of extreme demand, we will better understand which places benefit 
from greater economic resilience not only during the pandemic, but also into the 
future.

@@

...

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used 
interchangeably, as most remote labour during the Covid-19 crisis was carried out 
in the homes of individual employees rather than any other location [@eurofound2020]. 
However, it should be noted that previous research has explored how telecommuting 
can occur in other places, including satellite offices or on public transport 
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of 
telecommuting within different workforces, distinguising, for example, between those 
directly employed, indirectly employed, self-employed, full-time or part-time, and 
who use digital technologies to work remotely full-days or part-days 
[@allen2015effective; @bailey2002review; @haddad2009examination].
<!--old text: An appropriate definition of teleworking is ‘the remote provision of labour that would otherwise be carried out within company premises’ (European Commission, 2020b).-->
No matter the definition, the option and capability to telecommute or work from home 
has never been equally distributed spatially or socio-economically any more than 
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK indicate 
that telecommuters are most likely to hold professional, managerial, and technical 
occupations, where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].
Opportunities for working from home during the current pandemic have likewise not 
been equally spread across the workforce.
@NBERw26948 indicated that in the US, managers, educators, as well as those
working in computer-related occupations, finance, and law can easily work from
home, and that occupations with opportunities to telecommute are associated
with higher earnings.
This is not the case for the workforce occupied in more spatially fixed occupations,
from farming, construction, and manufacturing to hospitality and care services.
In the US, these occupations tend to be lower-income, non-white, 
without a university degree, live in rental accommodation and lack health insurance
[@NBERw27085].
Similar trends can be observed for other countries.
For example, $75$% of workers with tertiary education worked from home in Europe
during spring $2020$, whilst only $34$% of workers with secondary education and
$14$% of those primary education did so [@eurofound2020].

## digital divides and economic resilience
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be 
described as a manifestation of the third level digital divide, as those who are able 
to use digital technologies to work from home benefit from a high rate of return on 
their use of the internet in terms of autonomy, flexibility, and time saved from 
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis, when those with the 
ability to telecommute also have the ability to maintain their employment whilst 
protecting their health.
However, the success of these arrangements has also been dependent upon the first 
level digital divide, which is associated with access and quality of internet 
connectivity.
For example, the systematic review from @SALEMINK2017360 highlights the 
infrastructure quality differences between urban and rural areas.
Whether this variation in infrastructure quality affects the spatial footprint of 
telecommuting has not previously been measured, although there are indications that
those who purchase high speed connections consume more data of all sorts and use their
connections for a variety of purposes [@hauge2011consumer].
Another study identifies a correlation between access to internet services and a 
reduction in household transport spend across 33 coutries, although the authors are 
uncertain whether this implies less travel, for example because of increased
telecommuting, or whether internet access enables more efficient travel 
[@bris2017ict].
However, the extreme demand during the pandemic provides a new opportunity to 
understand how infrastructure accessibility, quality, and reliability affects 
telecommuting, particularly as working from home during the pandemic required
high volumes of bandwidth-intensive video conferencing in order to avoid the 
face-to-face contact that could increase the spread of infection.
Thus by answering questions about internet service resilience, we can also refine our
understanding of how this has contributed to or reduced the new digital divide where
economic resilience has been dependent upon the capability to work from home.

<!-- Econ resilience-->
This is aligned with the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic].

## other subsection?

- Introduce space-time geography
- circumstances where work can be carried out more flexibly in space and time
- how relate to digital divides
- capability theory of mobility? suppressed demand...

## covid and working from home, cities, urban structure
- economic geography of UK

@@

<!-- HB old PARA3 -->
This paper assesses the quality and reliability of upload speeds by time of day and 
day of the week across local authorities in the UK during the time when the population
were told to work from home if at all possible.
The experience of upload speeds is not the same as the maximum speeds offered by an 
ISP or possible speeds over a particular type and length of connection. 
We use volunteered geographic information in the form of speed-checks run by users to 
test their experienced broadband speeds, upload and download. 
Since the quality and reliability of internet services vary in time and space due to 
both supply and demand-side influences, we use a number of different measures of 
experienced upload speeds. 
These include: a) mean, experienced connection speed, 
b) standard deviation or the amount of fluctuation from the mean, and 
c) the variation in speeds at particular times of day when working from home is more 
likely to take place. 
We take account of all three measurements in order to describe upload speeds as fully 
as possible. 
We start by calculating the mean upload speed for every local government district in 
the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank 
holidays. 
We then cluster the local authorities by these temporal profiles, allowing us to 
identify patterns and describe the overall means, standard deviations, and other 
relevant statistics for each spatial cluster. 
Thus we aim to answer the first part of our research question: 
How resilient are broadband speeds as experienced in different parts of the UK during 
a time of extreme demand?

<!-- HB old PARA4 -->
The cause of these different experiences of broadband resilience may be different in 
different areas, as they may reflect similarities in patterns of demand or similar 
quality of infrastructure. 
Our approach is also limited by potential endogeneity, as for example, better quality 
connections with high mean speeds may enable more working from home, but greater 
demand may cause slower speeds, less reliability and greater variability of speed at 
different times of day or week. 
Therefore, we avoid attributing any cause to our analysis of the experienced level of 
quality and reliability of upload speeds. 
Instead, we run auxilliary regressions in order to understand how the spatial and 
temporal patterns of internet service relate to the economic geography of the UK. 
We discuss how the different patterns might support or undermine efforts to work from 
home and maintain safe productivity and whether they reinforce existing spatial and 
social inequalities. 
From this analysis, we hope to provide a greater understanding of how telecommuting 
and technology intersect at a time of extreme demand, and what lessons this time has 
for a future where telecommuting is likely to remain a common means of accessing work 
and broadband services, as well as infrastructure, must be fit for purpose.

The primary data analysed in this paper was provided by Speedchecker Ltd, a private company that allows internet users to check their own broadband upload / download speeds, and stores every speed-check with a timestamp and geolocation.

Our approach involves aggregating all the speed-checks during the 13 weeks of March to May inclusive for weekdays in 2020 by each hour of the day and day of the week. As our research aims to identify the geography of internet service resilience for work purposes, bank holidays and the hours between midnight and 6am were also excluded. The composite week time series thus comprise 18 hours multiplied by 5 weekdays or 90 time points per series. These time series were calculated for each of the 382 Local Authority Districts (LADs) in the UK, standardised, and then a k-means clustering algorithm was applied, including dynamic time warping. The LADs were assigned to 10 clusters for upload speeds. The cluster id for each LAD is then reattached to the speed-check dataframe to identify the characteristics of each cluster, including number of LADs, and the descriptives statistics of upload speeds in that cluster, and the temporal profile by hour of the day and day of the week.

@@

@@

Description of the method

@@

The temporal profiles used to cluster the local authorities have been summarised in graph [.], which shows a composite profile of mean upload speeds per hour per day for each cluster. For upload speeds, 345 of 382 local authorities, or over 62 million people, fall into cluster 6 or cluster 9. Graph [.] shows that both of these clusters have relatively similar temporal profiles, which are flatter than the other, smaller clusters, suggesting better reliability of service. However, the upload speeds at all times for cluster 9 are substantially higher than for cluster 6, which is an indication of better quality of service. This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59, as well as the evening peak period from 19:00-20:59. In comparison, the time profile in graph [.] shows upload speeds in cluster 1 are on average lower at certain times of day during the study period than any other cluster, whilst the profile for cluster 3 appears to show speeds fluctuating as much as cluster 1, but at levels usually higher even than cluster 9.

The variability of the smaller clusters may be related to the fewer speed tests from fewer local authorities that have been averaged, whilst averaging greater numbers of speed tests could artificially flatten the profile. This appears not to be the case for cluster 9, as there is negligible difference between morning and evening upload speeds, at 1% slower in the morning, confirming a high level of reliability of service. In comparison, upload speeds in cluster 6 are 4% slower in the morning than in the evening and experienced the joint highest ratio of standard deviation to mean across the time period under assessment. This suggests that although the time profile is relatively flat in graph [.], the experience is one of speeds that fluctuate from a lower mean, and therefore might more often impact on online activities. Still, in most of the smaller clusters the reliability of service is worse than in cluster 9 in terms of the ratio of standard deviation to mean and worse than cluster 6 as well in terms of the ratio of upload speed in the morning peak compared to the evening peak. In five of the smaller clusters, which are home to almost 3 million people, speeds are 12-19% slower between 9:00-10:59 compared to 19:00-20:59.

The quality and reliability of broadband service is thus much better in the 115 Local Authority areas in cluster 9, which are mostly in urban or suburban areas, compared to cluster 6. These include 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation. There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, and many suburban districts from the South East of England to South Tyneside. Meanwhile, the 230 local authorities in cluster 6, which have lower speeds on average and more variation in service still include major urban areas, such as Bristol, Liverpool and Leeds, and many suburban areas, but also include some of the most rural areas in the country. Meanwhile, Cluster 1, with 10 local authorities that are home to over 1 million people has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean. This cluster's most populous area is Westminster in central London.

@@

# Conclusions
<!-- WFH broader consequences-->
The long-term effects of such drastic changes in telecommuting and attitudes
towards working from home are difficult to predict. Nevertheless,
they span through various aspects of economy and society: from changes to
transportation planning due to altered commuting patters, to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713]; and from productivity and innovation changes, to changes
in agglomeration externalities and the attraction of large cities [@econobs]
just to name a few. This paper is positioned to support endeavours in
understanding the effects of increased telecommuting by
exposing the spatial and social dimensions of telecommuting including the
resilience of broadband speeds in terms of both quality and reliability of
service, and whether this reinforces or redresses prior digital divisions.
**HANNAH, HOW DOES THIS SOUND - this last bit probably in the discussion at the end?**

>>>>

During the pandemic, working from home using digital technologies, whether partially 
or exclusively, was transformed from a niche means of accessing work, albeit one that 
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs and,
to a varying extent, maintained productivity during periods of strict
lockdown around the world.
However, this ability has not been evenly distributed socially or spatially,
creating a new type of digital divide.
On one side are those who can work from home, supported by digital technologies, 
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this new digital divide can be viewed as mainly occupational, 
we consider whether the divide is also technological.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the resilience
of broadband speeds in terms of both quality and reliability of service may reinforce
or redress the spatial and social dimensions of the digital division exposed by the
pandemic.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK. **need to add urban-rural to analysis**
Our analysis enables us to better understand how the spatial and social distribution 
of both occupations and online accessibility intersect to enable or hinder the 
practice of telecommuting at a time of extreme demand. 
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a common means of accessing work, and broadband
services and infrastructure must be fit for purpose.
**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**

<!-- capability to WFH -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of 
location or co-location with colleagues, including supervisors 
[@allen2015effective; @singh2013modeling].
However, successful telecommuting also requires that the quality and
reliability of digital services, particularly home internet connection speeds,
enable the completion of work tasks with a minimum of delay or interruption.
Prior to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and connecting to
colleagues and workplace resources via broadband was the purview of a small minority 
of workers. 
Instead, leisure use in the evening, when video streaming services are at their peak, 
has been used to benchmark broadband performance and service delivery by different 
Internet Service Providers (ISPs), at least in the UK [@ofcom2017].
Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than pre-Covid, raising questions around whether internet 
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked soley from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020; 
@ons2020lm2019].
A back of the envelope calculation suggests that up to 40% of the working force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries [@felstead2020homeworking].
For instance, $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid], 
and a recent estimate indicated that $37$% of all jobs in the US can be permanently 
performed entirely from home [@NBERw26948].

<!-- ICT infrastructure and digital divides-->
None of these changes could have happened in the absence of reliable ICT
infrastructure -- both in terms of software and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature describes first level digital divides in terms of the availability and
quality of internet connectivity, such as that manifest in different geographies 
in the UK [@riddlesden2014broadband; @philip2017digital]. 
Second level digital divides consider the presence or lack of the necessary skills 
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the differentiated returns of internet use **for...? add a phrase here introducing economic resilience?**
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic 
resilience of people and places during the current pandemic.
We identify this as a new digital divide, one that fundamentally alters the 
potential returns of internet use for the user and wider community, 
assumes skills or functions that are present in some occupations but not in others, 
and relies upon access to high quality internet services.
As the quality of internet infrastructure and services, as well as variation in 
occupations are spatially dependent and clustered in space, our approach offers a 
framework for understanding which types of places,  
are more likely to land on the right side of this new digital divide. 
By asking how resilient broadband speeds are as experienced in different parts of the 
UK during a time of extreme demand, 
we interrogate which places benefit from greater the economic resilience 
digital technologies can offer, not only during the pandemic, 
but also into the future.

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used 
interchangeably, as most remote labour during the Covid-19 crisis was carried out 
in the homes of individual employees rather than any other location [@eurofound2020]. 
However, it should be noted that previous research has explored how telecommuting 
can occur in other places, including satellite offices or on public transport 
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of 
telecommuting within different workforces, distinguishing, for example, between those 
directly employed, indirectly employed, self-employed, full-time or part-time, and 
those who use digital technologies to work remotely full-days or part-days 
[@allen2015effective; @bailey2002review; @haddad2009examination].
No matter the definition, the option and capability to telecommute or work from home 
has never been equally distributed spatially or socio-economically any more than 
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK indicate 
that telecommuters are most likely to hold professional, managerial, and technical 
occupations, where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].
Opportunities for working from home during the current pandemic have likewise not 
been equally spread across the workforce.
@NBERw26948 indicated that in the US, managers, educators, as well as those
working in computer-related occupations, finance, and law can easily work from home,
and that occupations with opportunities to telecommute are associated
with higher earnings.
This is not the case for the workforce occupied in more spatially fixed occupations,
from farming, construction and manufacturing to hospitality and care services.
In the US, these occupations tend to be lower-income, non-white, 
without a university degree, live in rental accommodation and lack health insurance
[@NBERw27085].
Similar trends can be observed for other countries.
For example, $75$% of workers with tertiary education worked from home in Europe
during spring $2020$, whilst only $34$% of workers with secondary education and
$14$% of those primary education did so [@eurofound2020].

## digital divides and economic resilience
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be 
described as a manifestation of the third level digital divide, as those who are able 
to use digital technologies to work from home benefit from a high rate of return on 
their use of the internet in terms of autonomy, flexibility, and time saved from 
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis, 
when those with the ability to telecommute also have the ability to maintain 
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first 
level digital divide, which is associated with access and quality of internet 
connectivity.
For example, the systematic review from @SALEMINK2017360 highlights the 
infrastructure quality differences between urban and rural areas 
in various advanced economies.
Whether this variation in infrastructure quality affects the spatial footprint of 
telecommuting has not previously been measured, although there are indications that
those who purchase high speed connections consume more data of all sorts and use their
connections for a variety of purposes [@hauge2011consumer].
Another study identifies a correlation between access to internet services and a 
reduction in household transport spend across 33 coutries [@bris2017ict]. 
Whether the implication is less travel because of increased telecommuting, 
or whether internet access enables more efficient travel, this finding is 
an indication of the potential household savings better internet services offer.
Thus, the extreme demand during the pandemic provides a new opportunity to 
understand how infrastructure accessibility, quality, and reliability 
affects telecommuting, particularly as working from home during the pandemic required
high volumes of bandwidth-intensive video conferencing in order to avoid the 
face-to-face contact that could increase the spread of infection.
By first answering questions about internet service resilience, we can also refine our
understanding of how this has contributed to or reduced the new digital divide, 
where economic resilience has been dependent upon the capability to work from home.

## covid and working from home, cities, urban structure
This is aligned with the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic]. 
Different cities have different industrial and occupational profiles, 
and these will affect the potential for telecommuting... 
**Fill in more here - relate to 2nd and 3rd level digital divide, economic geography of UK**

## other subsection?
- circumstances where work can be carried out more flexibly in space and time
- how relate to digital divides
- capability theory of mobility?
- broadband tech stuff? broadband studies / divides / resilience

@@

@@

<!-- PARA introduce broadband speeds -->
This paper assesses the quality and reliability of upload speeds by time of day and 
day of the week across local authorities in the UK during the time when the population
were told to work from home if at all possible.
The experience of upload speeds is not the same as the maximum speeds offered by an 
ISP or possible speeds over a particular type and length of connection. 
We use volunteered geographic information in the form of speed-checks run by users to 
test their experienced broadband speeds, upload and download. 
This primary data was provided by Speedchecker Ltd, 
a private company that allows internet users to check their own broadband upload / 
download speeds, and stores every speed-check with a timestamp and geolocation.
Since the quality and reliability of internet services vary in time and space due to 
both supply and demand-side influences, we use a number of different measures of 
experienced upload speeds. 
These include: a) mean, experienced connection speed, 
b) standard deviation or the amount of fluctuation from the mean, and 
c) the variation in speeds at particular times of day when working from home is more 
likely to take place. 
We take account of all three measurements in order to describe upload speeds as fully 
as possible.
The following **graphs / tables** show these statistics for the sample as a whole,
as well as similar statistics related to frequency of testing.

@@

**Just looking back, it was partly the frequency of tests, but that was also by geography, so do we need maps? And I feel there was something else I'm forgetting...**

@@

Clustering methods are a widely used family of techniques in **economic?** geography (**ADD REF**).
It can be defined within the modern machine learning framework as an unsupervised
learning task, which involves partitioning unlabeled observations into homogeneous
groups known as clusters [@montero2014tsclust]. 
The key idea is that observations within clusters tend to be more similar than 
observations between clusters.
Clustering is particularly useful for exploratory studies as it identifies
structures within the data [@aghabozorgi2015time].

Various applications of clustering can be found within geography: **ADD EXAMPLES**

Key characteristic of these studies is the cross-sectional nature of the data they
employed. 
Indeed, most clustering problems in geography deal with observations 
that are fixed in time. 
However, for this paper we are interested in creating clusters of local authorities 
in the UK with similar temporal signatures of experienced internet speeds over time. 
Hence, we deviate from the established geographical clustering tools and 
employ time series clustering methods.

Time-series is a common type of dynamic data that naturally arises in many 
different scenarios, such as stock data, medical data, and machine monitoring, 
just to name a few **(Aghabozorgi et al., 2015; Aggarwal and Reddy, 2013)**. 
They pose some challenging issues due to the large size and high dimensionality 
commonly associated with time-series **(Aghabozorgi et al., 2015)**. 
In this context, dimensionality of a series is related to time, 
and it can be understood as the length of the series.

Time series clustering methods have been developed in order to deal with clustering
problems linked to, for instance, stock or other financial data, economic,
governmental or medical data as well as machine monitoring
[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
The main challenge -- and also the difference with cross-sectional clustering
problems -- is data dimensionality given the multiplicity of data points for every
individual included in the data set. 
Time series are dynamic data as the value of the observations change as 
a function of time [@aghabozorgi2015time].
This high dimensionality leads to 
(i) computational and algorithmic challenges regarding handling these data and 
building algorithms to perform clustering over long time series, 
and (ii) open questions regarding the choice of similarity measures in order to 
cluster similar times series together considering the whole length of the time series 
and overcoming issues around noise, outliers and shifts [@lin2004iterative; 
@aghabozorgi2015time].

Time series clustering methods utilising the whole length of time series can be
grouped in three categories. 
The first -- model-based approaches -- is based on recovering the underlying model 
for each time series and then apply clustering algorithms on the model parameters 
[@aghabozorgi2015time]. 
The main criticism is the cluster accuracy for near clusters [@mitsa2010temporal]. 
The second approach is a based the formation of vectors of features based on the 
original time series.
These new data of reduced dimensionality is then clustered using conventional
clustering algorithms.

For this paper we utilise the third category of time series clustering methods
known as shape-based approaches. 
The main idea is to match any two sperate time series based on the similarity 
of their shapes through the calculation of distances among these two shapes.

@sardatime
Ideally, all members of the same cluster are similar to each other, but are as
dissimilar as possible from objects in a different cluster.

the data and how the underlying grouping is performed. 
One classification depends on whether the whole series, a subsequence, 
or individual time points are to be clustered. 
On the other hand, the clustering itself may be shape-based, feature-based, or model-based.

In the context of shape-based time-series clustering, it is common to utilize the Dynamic Time Warping (DTW) distance as dissimilarity measure **(Aghabozorgi et al., 2015)**. 
The calculation of the DTW distance involves a dynamic programming algorithm 
that tries to find the optimum warping path between two series under certain constraints

**moved this section from data and descriptives as need to end methods of clustering with how it applies to what we did, but needs further edit after you're done**
To apply these methods to our data, 
we aggregate all the speed-checks during the 13 weeks of March 
to May inclusive for weekdays in 2020 by each hour of the day and day of the week. 
As our research aims to identify the geography of internet service resilience for 
work purposes, bank holidays and the hours between midnight and 6am were also 
excluded. 
The composite week time series thus comprise 18 hours multiplied by 5 weekdays or 
90 time points per series. 
These time series were calculated for each of 
the 382 Local Authority Districts (LADs) in the UK, standardised, 
and then a k-means clustering algorithm was applied, including dynamic time warping. 
The LADs were assigned to 10 clusters for upload speeds. 
The cluster id for each LAD is then reattached to the speed-check dataframe to 
identify the characteristics of each cluster, including number of LADs, and the 
descriptives statistics of upload speeds in that cluster, 
and the temporal profile by hour of the day and day of the week.
Thus we aim to answer the first part of our research question: 
How resilient are broadband speeds as experienced in different parts of the UK during 
a time of extreme demand?

<!-- HB old PARA4 - link to auxillary regressions method, or put in results to introduce results of auxillary regressions? -->
The cause of these different experiences of broadband resilience may be different in 
different areas, as they may reflect either similarities in patterns of demand 
or similar quality of infrastructure. 
Our approach is also limited by potential endogeneity, as for example, better quality 
connections with high mean speeds may enable more working from home, but greater 
demand may cause slower speeds, less reliability and greater variability of speed at 
different times of day or week. 
Therefore, we avoid attributing any cause to our analysis of the experienced level of 
quality and reliability of upload speeds. 
Instead, we run auxilliary regressions in order to understand how the spatial and 
temporal patterns of internet service relate to the economic geography of the UK. 
We discuss how the different patterns might support or undermine efforts to work from 
home and maintain safe productivity and whether they reinforce existing spatial and 
social inequalities. 
From this analysis, we hope to provide a greater understanding of how telecommuting 
and technology intersect at a time of extreme demand, and what lessons this time has 
for a future where telecommuting is likely to remain a common means of accessing work 
and broadband services, as well as infrastructure, must be fit for purpose.

@@

The temporal profiles used to cluster the local authorities have been summarised in 
graph [.], which 
shows a composite profile of mean upload speeds per hour per day for each cluster. 
For upload speeds, 345 of 382 local authorities, or over 62 million people, 
fall into cluster 6 or cluster 9. 
Graph [.] suggests that these two clusters have relatively similar temporal profiles, 
which are flatter than the other, smaller clusters, 
suggesting better reliability of service. 
However, the upload speeds at all times for cluster 9 are substantially higher 
than for cluster 6, which is an indication of better quality of service. 
This difference is reflected not only in the mean speeds for these clusters for the 
whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59, 
as well as the evening peak period from 19:00-20:59. 
In comparison, the time profile in graph [.] shows upload speeds in cluster 1 are 
on average lower at certain times of day during the study period than any other 
cluster, whilst the profile for cluster 3 appears to show speeds fluctuating as much 
as cluster 1, but at levels usually higher even than cluster 9.

The variability of the smaller clusters may be related to the fewer speed tests from 
fewer local authorities that have been averaged, whilst averaging greater numbers of 
speed tests could artificially flatten the profile. 
This appears not to be the case for cluster 9, as there is negligible difference 
between morning and evening upload speeds, at 1% slower in the morning, confirming a 
high level of reliability of service. 
In comparison, upload speeds in cluster 6 are 4% slower in the morning than in the 
evening and experienced the joint highest ratio of standard deviation to mean across 
the time period under assessment. 
This suggests that although the time profile is relatively flat in graph [.], 
the experience is one of speeds that fluctuate from a lower mean, 
and therefore might more often impact on online activities. 
Still, in most of the smaller clusters the reliability of service is worse than in 
cluster 9 in terms of the ratio of standard deviation to mean. 
The smaller clusters are mainly worse than clusters 9 and 6 in terms of 
the ratio of upload speed in the morning peak compared to the evening peak. 
In five of the smaller clusters, which are home to almost 3 million people, 
speeds are 12-19% slower between 9:00-10:59 compared to 19:00-20:59.

The quality and reliability of broadband service is much better in 
the 115 Local Authority areas in cluster 9, 
which are mostly in urban or suburban areas, compared to cluster 6. 
These include 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation. There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, and many suburban districts from the South East of England to South Tyneside. Meanwhile, the 230 local authorities in cluster 6, which have lower speeds on average and more variation in service still include major urban areas, such as Bristol, Liverpool and Leeds, and many suburban areas, but also include some of the most rural areas in the country. Meanwhile, Cluster 1, with 10 local authorities that are home to over 1 million people has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean. This cluster's most populous area is Westminster in central London.

@@

# Conclusions

<!-- WFH broader consequences-->
The long-term effects of such drastic changes in telecommuting and attitudes
towards working from home are difficult to predict. Nevertheless,
they span through various aspects of economy and society: from changes to
transportation planning due to altered commuting patters, to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713]**also 2020 Swedish article from JTG**; and from productivity and innovation changes, to changes
in agglomeration externalities and the attraction of large cities [@econobs]
just to name a few. This paper is positioned to support endeavours in
understanding the effects of increased telecommuting by
exposing the spatial and social dimensions of telecommuting including the
resilience of broadband speeds in terms of both quality and reliability of
service, and whether this reinforces or redresses prior digital divisions.
**took this bit you wrote to put in the discussion at the end?**
